From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-07-30 WOLFHPC15/article/concept.tex
fix 1,2,3,4 helene 30_07

%----------------------------------------
\subsection{Stencil codes}
\label{sect:stencil}
%----------------------------------------
To numerically solve a set of PDEs, iterative methods (finite difference, finite volume, finite element methods etc.) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, in a mesh-based numerical simulation, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain (the mesh). Among those numerical computations is found a set of numerical schemes, also called \textit{stencil computations}~\cite{spaaTangCKLL11}. To ease programming of stencil computations, and because of the relative regularity of such a computation, many solutions (languages or libraries) propose an easy way to describe the stencil code, while producing transparently an optimized and parallelized stencil code. For example, Liszt~\cite{DeVito2011LDS} and OP2~\cite{Giles2011} are domain specific languages to define a stencil code on unstructured meshes, while Pochoir~\cite{spaaTangCKLL11} and PATUS~\cite{citeulike12258902} address structured meshes, respectively proposing an optimized cache tiling technique, and a separation of concerns between parallelization strategies and computation descriptions.

\fix{cp2hc: transition a revoir.}

Let introduce some concepts used in this paper. First, a stencil code is based on a \emph{mesh} which is the discretization of the physical real domain. A mesh is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). For example, a structured mesh is illustrated in Figure~\ref{fig:mesh}. Each cell contains four nodes and four edges. Second, \emph{mesh domains} are subset elements of the mesh, such as the center of cells, edges or nodes. In Figure~\ref{fig:mesh}, two mesh domains are illustrated, the center of the cells (named \texttt{Cells}, in red) and edges on the horizontal axe (named \texttt{Edgex}, in blue on left and right of each cell). A \emph{data} is a quantity to simulate or a temporary data to perform computations. A data is mapped onto a mesh domain. For example, in Figure~\ref{fig:ex1}, data $A$ and $B$ are mapped onto the center of cells, while, in Figure~\ref{fig:ex2}, $C$ is mapped onto the edges.

@@

The stencil computation in Figure~\ref{fig:ex1} computes $A$ using $B$, while the one in Figure~\ref{fig:ex2} computes $A$ using $C$. Thus, a stencil computation is defined by a set of input data (in those examples only one input data), and a single output data, the result. The \emph{computation domain} is a subset of elements to compute on the output data. In other words a computation domain is a subset of a mesh domain. For example, in the first stencil, the computation of $A$ is performed on plain elements, while dotted elements are not computed. On the other hand, on the second computation, the computation domain of $A$ is exactly equal to the cell mesh domain. The computation domain represents the set of elements on which the computation is performed, in other words, the space loop.

Finally, a stencil computation is composed of a \emph{numerical expression} which indicates how to compute an element of the computation domain (on the output data), using elements of the set of input data. The numerical expression of a stencil computation has the particularity to use not only the element to compute but also a \emph{neighborhood} of this element, also called a \emph{stencil shape}. For example, the stencil shape of the first computation in Figure~\ref{fig:ex1} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood shape can also access a different mesh domain, as for example in Figure~\ref{fig:ex2}. Actually, in this computation, the neighborhood contains edges on the left and on the right of a cell. As an example, the numerical expression of the first example could be:
\begin{equation*} 
A(x,y) = B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1),
\end{equation*}
and the numerical expression of the second example could be:
\begin{equation*} 
A(x,y) = A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

%----------------------------------------
\subsection{A Multi-Stencil Formlization}
\label{sect:multistencil}
%----------------------------------------
A real case numerical simulation is most of the time composed of more than one stencil computation, with one or more stencil shapes, and onto one or more data, and of additionnal local computations.
\fix{cp2hc: [optionnelle] Par example? domaine applicatif?}
Let introduce some formal definitions used in the rest of the paper.

For $\Delta$ the set of data of the simulation, a stencil computation $s$ is defined as the quintuplet
\begin{equation} 
s(R,w,exp,d_c),
\label{eq:st}
\end{equation}
where $R$ is a set of pair $(r,n)$, with $r \in \Delta$ is a data read by the computation, and $n$ is a stencil shape (a neighborhood) applied on $r$. The data written by the computation is denoted $w \in \Delta$, $exp$ is the numerical expression of the stencil computation. Finally the numerical expression is applied on the computation domain $d_c$.

For example, in the example of Figure~\ref{fig:ex1}, assuming the computation domain (plain lines) is $dc1$ and the stencil shape is $n1$, the stencil computation can be defined with:
\begin{equation*}
R: {(B,n1)} \quad w: A \quad d_c: dc1
\end{equation*}
\begin{equation*}
exp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil computation can be defined with:
\begin{equation*}
R: {(C,n2),(A,null)} \quad w: A \quad d_c: dc2
\end{equation*}
\begin{equation*}
exp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1)
\end{equation*}
One can notice that the input data $A$ is associated to the stencil shape $null$. This means that no stencil shape is applied on $A$, but only local accesses.

A local, or auxiliary, computation is almost the same than a stencil computation without neighborhood information in the numerical expression, which makes this kind of computation purely local. A local computation $l$ is defined as the quadruplet:
\begin{equation} 
l(R_l,w,exp,d_c),
\label{eq:loc}
\end{equation}
where $R_l$ is the set of input data, and other parameters are the same than for a stencil computation.

A multi-stencil program in this work is defined as a sextuplet:
\begin{equation} 
\mathcal{MSP}(T,\mathcal{M},\mathcal{D}_m,\mathcal{D}_c,\Delta,\Gamma),
\label{eq:msp}
\end{equation}
where $T$ is the set of time iteration to run the simulation, $\mathcal{M}$ is the mesh of the simulation, $\mathcal{D}_m$ is the set of mesh domains, $\mathcal{D}_c$ is the set of computation domains used into computations, $\Delta$ is the set of data of the simulation, each one mapped onto a mesh domain, and finally $\Gamma$ is the set of computations. The set of computations $\Gamma$ is composed of an ordered list of stencil and local computations. One can notice that this work is limited to a single type of mesh in a given simulation.

>>>>

%----------------------------------------
\subsection{Stencil codes}
\label{sect:stencil}
%----------------------------------------
To numerically solve a set of PDEs, iterative methods (finite difference, finite volume, finite element methods etc.) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, in a mesh-based numerical simulation, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain (the mesh). Among those numerical computations is found a set of numerical schemes, also called \textit{stencil kernels}~\cite{spaaTangCKLL11}. %To ease programming of stencil kernels, and because of the relative regularity of such a computation, many solutions (languages or libraries) propose an easy way to describe the stencil code, while producing transparently an optimized and parallelized stencil kernel. For example, Liszt~\cite{DeVito2011LDS} and OP2~\cite{Giles2011} are domain specific languages to define a stencil kernel on unstructured meshes, while Pochoir~\cite{spaaTangCKLL11} and PATUS~\cite{citeulike12258902} address structured meshes, respectively proposing an optimized cache tiling technique, and a separation of concerns between parallelization strategies and computation descriptions.

%\fix{cp2hc: transition a revoir.}

In this paper are used rather unusual concepts to define a stencil, let introduce those concepts. First, a stencil kernel is based on a \emph{mesh} which is the discretization of the physical real domain. A mesh is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). For example, a structured mesh is illustrated in Figure~\ref{fig:mesh}. Each cell contains four nodes and four edges. Second, \emph{mesh entities} are subset elements of the mesh, such as the center of cells, edges or nodes. In Figure~\ref{fig:mesh}, two mesh entities are illustrated, the center of the cells (named \texttt{Cells}, in red) and edges on the horizontal axe (named \texttt{Edgex}, in blue on left and right of each cell). A \emph{data} is a quantity to simulate or a temporary data to perform computations. A data is mapped onto a given mesh entity. For example, in Figure~\ref{fig:ex1}, data $A$ and $B$ are mapped onto the center of cells, while, in Figure~\ref{fig:ex2}, $C$ is mapped onto the edges.

@@

The stencil kernel of Figure~\ref{fig:ex1} computes $A$ using $B$, while the one of Figure~\ref{fig:ex2} computes $A$ using $C$. Thus, a stencil kernel is defined by a set of input data (in those examples only one input data), and a single output data, the result. The \emph{computation domain} is a subset of elements to compute on the output data. In other words a computation domain is a subset of a mesh entity. For example, in the first stencil kernel, the computation of $A$ is performed on elements represented with full lines, while dotted elements are not computed. On the other hand, on the second kernel, the computation domain of $A$ is exactly equal to the cell mesh entity. The computation domain represents the set of elements on which the kernel is performed, in other words, the space loop.

Finally, a stencil kernel is composed of a \emph{numerical expression} which indicates how to compute an element of the computation domain (on the output data), using elements of the set of input data. The numerical expression of a stencil kernel has the particularity to use not only the element to compute but also a \emph{neighborhood} of this element, also called a \emph{stencil shape}. For example, the stencil shape of the first computation in Figure~\ref{fig:ex1} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood shape can also access a different mesh entity, as for example in Figure~\ref{fig:ex2}. Actually, in this second kernel, the neighborhood contains edges on the left and on the right of a cell. As an example, the numerical expression of the first kernel could be:
\begin{equation*} 
A(x,y) = B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1),
\end{equation*}
and the numerical expression of the second kernel could be:
\begin{equation*} 
A(x,y) = A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

To resume, concepts used in this paper to define a stencil kernel are \emph{mesh}, \emph{mesh entity}, \emph{data}, \emph{computation domain}, \emph{numerical expression} and \emph{stencil shape}. 
If stencil kernels have been studied a lot, the formalization of real overall mesh-based numerical simulations is a rather new contribution. Actually, paying attention to complex numerical simulations, it appears that most of them are composed of more than one stencil kernel, with one or more stencil shapes and of additionnal local computations. For example, we would like to formalize and parallelize a numerical simulation which chains stencil kernels of Figures~\ref{fig:ex1} and~\ref{fig:ex2}. The next section formalize concepts of \emph{stencil kernel}, \emph{local computation} and what we call a \emph{multi-stencil program}.
%\fix{cp2hc: [optionnelle] Par example? domaine applicatif?}

%----------------------------------------
\subsection{A Multi-Stencil Formalization}
\label{sect:multistencil}
%----------------------------------------
Let introduce some formal definitions used in the rest of the paper.
For $\Delta$ the set of data of the simulation, a stencil kernel $s$ is defined as the quintuplet
\begin{equation} 
s(R,w,exp,d),
\label{eq:st}
\end{equation}
where $R$ is a set of pair $(r,n)$, with $r \in \Delta$ is a data read by the computation, and $n$ is a stencil shape (a neighborhood) applied on $r$. The data written by the kernel is denoted $w \in \Delta$, $exp$ is the numerical expression of the stencil kernel. Finally the numerical expression is applied on the computation domain $d$.

For example, in Figure~\ref{fig:ex1}, assuming the computation domain (full lines) is $dc1$ and the stencil shape is $n1$, the stencil kernel can be defined with:
\begin{equation*}
R: {(B,n1)} \quad w: A \quad d: dc1
\end{equation*}
\begin{equation*}
exp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil kernel can be defined with:
\begin{equation*}
R: {(C,n2),(A,null)} \quad w: A \quad d: dc2
\end{equation*}
\begin{equation*}
exp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1)
\end{equation*}
One can notice that the input data $A$ is associated to the stencil shape $null$. This means that no stencil shape is applied on $A$, but only local accesses.

A local, or auxiliary, computation is a particular case of stencil kernel without neighborhood information in the numerical expression, which makes this kind of computation purely local. A local computation $l$ is defined as the quadruplet:
\begin{equation} 
l(R_l,w,exp,d),
\label{eq:loc}
\end{equation}
where $R_l$ is the set of input data, and other parameters are the same than for a stencil computation.

A multi-stencil program in this work is defined as a sextuplet:
\begin{equation} 
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma),
\label{eq:msp}
\end{equation}
where $T$ is the set of time iteration to run the simulation, $\mathcal{M}$ is the mesh of the simulation, $\mathcal{E}$ is the set of mesh entities, $\mathcal{D}$ is the set of computation domains used into computations, $\Delta$ is the set of data of the simulation, each one mapped onto a mesh entity, and finally $\Gamma$ is the set of computations. The set of computations $\Gamma$ is an ordered list of stencil kernels and local computations.

\fix{HC 2 all : Est-ce que je laisse cette phrase : One can notice that this work is limited to a single type of mesh in a given simulation.}

### 2015-07-30 WOLFHPC15/article/introduction.tex
fix 1,2,3,4 helene 30_07

Domain Specific Languages (DSLs) are a promising solution to hide intricate details of HPC programming from non experts.
By being specific to a well-defined set of problems, these languages enable the generation of well parallelized and optimized code for various architectures while requiring only a limited amount of information from the developer; part of the knowledge is directly embedded in the compiler.
A critical question however concerns the specificity of DSLs.
The less specific the DSL is, the more informations its users must provide and the less automatic optimizations can be embedded in the compiler.
On the other side of the scope, the more specific the DSL is, the less different code can use it and the more difficult it is to amortize the implementation cost of the DSL (language, compiler, and runtime).
Programming difficulties, as well as non-maintainability and non-portability often seem to be simply deported from the scientist to DSL designers.
% Moreover, one can consider that end users need as much DSLs as domains, which keeps the amount of work to write and maintain up-to-date DSLs out of hands.

For this reason, solutions to ease the design and implementation of DSLs have recently appeared~\cite{Fernandez:2014:DFL:2691166.2691168}. The main idea of those solutions is to propose a single DSL to write new DSLs. In this paper, we study a first contribution to another approach to improve maintainability and code-reuse in DSLs. This approach proposes a transformation from a DSL to a component-based runtime.
To evaluate this approach, a proper new DSL for \emph{multi-stencil} programs, and its compiler are presented in this paper.
\fix{cp2cp+hc: to dev benefits of components.}

Many solutions to automatically optimize and parallelize partial differential equations (PDEs) solvers have been proposed in the literature, either as specific libraries~\cite{petsc-efficient,Trilinos-Overview,CPE:CPE3494} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil.
However, a real case numerical simulation is most of the time not composed of a single stencil computation, but of a set of stencil computations and a set of additional local auxiliary computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil program}. As it will be detailed in the related work, this is complementary to the optimization and parallelization of a single stencil computation.
\fix{JB: j'utiliserais bien le terme "kernel" ou "stencil kernel" pour ce que tu apelles "stencil" et "stencil" tout court pour ce que tu apelles "stencil shape"}

>>>>

Domain Specific Languages (DSLs) are a promising solution to hide intricate details of HPC programming from non experts.
By being specific to a well-defined set of problems, these languages enable the generation of well parallelized and optimized code for various architectures while requiring only a limited amount of information from the developer; part of the knowledge is directly embedded in the compiler.
A critical question however concerns the specificity of DSLs.
Actually, the less specific the DSL is, the more information its users must provide and the less automatic optimizations can be embedded in the compiler.
%On the other side of the scope
Conversely, the more specific the DSL is, the less number of different applications can use it and the more difficult it is to amortize the implementation cost of the DSL (language, compiler, and runtime).
Programming difficulties, as well as non-maintainability and non-portability often seem to be simply deported from the scientist to DSL designers.
% Moreover, one can consider that end users need as much DSLs as domains, which keeps the amount of work to write and maintain up-to-date DSLs out of hands.

For this reason, solutions to ease the design and implementation of DSLs have recently appeared~\cite{Fernandez:2014:DFL:2691166.2691168}. The main idea of those solutions is to propose a single DSL to write new DSLs. In this paper, we study a first contribution to another approach to improve maintainability and portability in DSLs. This approach proposes a transformation from a DSL to a component-based runtime. Component-based Software Engineering have proved many times good properties for code re-use, separation of concerns, maintainability and productivity of codes. For this reason, we think that mixing DSLs and component-based runtimes could enable those properties, by inheritance, in DSLs.
To evaluate this approach, a proper new DSL for \emph{multi-stencil} programs, and its compiler are presented in this paper.
%\fix{cp2cp+hc: to dev benefits of components.}

Many solutions to automatically optimize and parallelize partial differential equations (PDEs) solvers have been proposed in the literature, either as specific libraries~\cite{petsc-efficient,Trilinos-Overview,CPE:CPE3494} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil program}. As it will be detailed in the related work, this is complementary to the optimization and parallelization of a single stencil kernel.
%\fix{JB: j'utiliserais bien le terme "kernel" ou "stencil kernel" pour ce que tu apelles "stencil" et "stencil" tout court pour ce que tu apelles "stencil shape"}

### 2015-07-30 WOLFHPC15/article/ms_msc.tex
fix 1,2,3,4 helene 30_07

As mentionned in the previous section, the Multi Stencil Language (MSL) is an agnostic descriptive language for multi-stencil simulations. Six main sections are required in a MSL description, as the six tuples of the formal definition of a multi-stencil program: the mesh description; the mesh entities description; the computation domains description and their dependencies; the data description; the time loop description; and finally, the computations description. Figure~\ref{fig:grammar} shows the grammar of MSL. Lines 1 to 9 define what is a MSL program with the different parts memntionned above. The remaining of this section describes the parts of a MSL program and presents an example.
We can note that, as MSL is a descriptive language, almost all terminals $meshid$, $meshdom$, $compdomid$, $dataid$, $compid$ ans $neighborid$ are string identifiers. Only $num$ is an integer terminal.

\subsubsection*{Mesh and mesh entities}
As illustrated in the first line of the grammar in Figure~\ref{fig:grammar}, a mesh is directly defined by a string identifier. Mesh entities are described as a list of identifiers (lines 2, 10 and 11 in Figure~\ref{fig:grammar}). In the example given in Figure~\ref{fig:mslex} a Cartesian mesh, called $cart$, and three mesh entities, called $cell$, $edgex$ and $edgey$, are defined.

\subsubsection*{Computation domains and their dependencies}
A computation domain can be seen as a subpart of a mesh entity which will be used during one or more computations. For each computation domain, the mesh entity from which it is built is indicated (lines 3-4 and 12-14 in Figure~\ref{fig:grammar}). By default two computation domains always intersect except when an independency relation is mentionned in the section \emph{independent} of the language (lines 5-6 and 15-17 in Figure~\ref{fig:grammar}). In the example of Figure~\ref{fig:mslex}, on lines 4 to 6, an whole computation domain is defined for each mesh entity, called $allcell$, $alledgex$ and $alledgey$, and on lines 7 and 8 two subparts of the mesh entity $edgex$ are also defined, called $part1edgex$ and $part2edgex$. Line 10 means that $part1edgex$ and $part2edgex$ do not intersect.

\subsubsection*{Data and time}
A data is a quantity to simulate and is mapped onto a mesh entity, as illustrated in lines 7 and 18-20 of Figure~\ref{fig:grammar}. The $time$ section simplies indicates a number of iterations to perform in the simulation. In the example of Figure~\ref{fig:mslex} are defined ten data and 500 iterations (lines 11-22). An extension could be to support a convergence criterion.

\subsubsection*{Computations description}
Finally, the language follows the two definitions of a stencil computation and a local computation given in Section~\ref{sect:multistencil}. However as MSL does not handle numerical expressions, the term $exp$ of Definition~\ref{eq:st} and Definition~\ref{eq:loc} is not used in the language. The exact code of the computation as to be writtenin the final ready-to-fill parallel pattern of the simulation, using external languages or interfaces (\cf Section~\ref{sect:component}). Moreover, to propose an intuitive syntax, information are given in a different order than in formal definitions.
Thus, the description of a computation, as indicated in lines 24 and 25 of Figure~\ref{fig:grammar}, consists in first the identifier of the data to compute ($w$ in the formal description), in brackets the identifier of the computation domain on which the computation is applied ($d_c$ in the formal definition), an identifier for the computation, and the set of identifiers for the data read by the computation with parenthesis. For each data read by the computation brackets indicate the stencil shape identifier used on the data ($R$ in the formal definition of a stencil). If the computation is local no brackets appear ($R_l$ in the formal definition). In the example of Figure~\ref{fig:mslex}, nine computations are described. For example, on line 24, the data $b$ is computed on the computation domain $allcell$ by the computation $c_0$ which read data $a$ without neighborhood shape.

@@

%----------------------------------------
\subsection{MSC transformation}
%----------------------------------------
The MSC transformation, which is a subpart of the overall compiler MSCAC, makes use of the ordered list of computations $\Gamma$, which can directly been extracted from the parser, to build a parallel representation of the computations of the overall multi-stencil program. This transformation phase of the compilation is divided in four different steps; it transforms $\Gamma$ to a series-parallel tree decomposition~\ref{???}. This transformation phase handles data parallelism through the first step, and task parallelism through the three next steps. This section gives an overview of these four steps of MSC; the detailed formalism and associated algorithms of those steps are described in the research report~\cite{????}.

\subsubsection*{The Computation Ordered List: $\Gamma$}
$\Gamma$ is directly obtained from the parser. Actually, the list of computations in the MSL program is already ordered: $\Gamma$ is a direct map of this list. In the example of Figure~\ref{fig:mslex}, $\Gamma = [c_0,c_1,c_2,c_3,c_4,c_5,c_6,c_7,c_8]$.

\subsubsection*{The Computation Synchronized Ordered List: $\Gamma_{data}$}
Considering a data parallelization of the simulation, the mesh and data of the simulation are splitted among available processors (or cores) and a single program is applied on each sub-part with additionnal synchronizations between processors or cores. For example, as a stencil code access neighborhood value, values computed by another processor are needed and synchronized. 
From the ordered list of computations $\Gamma$, the needed synchronizations can be automatically detected. A synchronization is needed each time a data read by a stencil computation has been written by a previous computation. In such a case, a computation is added before the stencil computation. This \emph{synchronization computation} read the data to synchronize, and write the same data. The computation domain of such a synchronization computation is the mesh entity on which the data is declared. As a result $\Gamma$ is transformed to a synchronized ordered list of computations $\Gamma_{data}$.
For example, in Figure~\ref{fig:mslex}, the stencil computation $c_1$ read the data $b$ which has been written by the computation $c_0$. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed to the sublist $[c_0,sync_1,c_1]$ in $\Gamma_{data}$. The new computation $sync_1$ read and write $b$ and is applied on the computation domain of $c_1$. As a result, a dependency is kept between $c_0$ and $sync_1$ and between $sync_1$, and $c_1$. The same transformation is performed for the stencils $c_4$ and $c_8$. Thus, $\Gamma_{data} = [c_0,sync_1,c_1,c_2,c_3,sync_4,c_4,c_5,c_6,c_7,sync_8,c_8]$.

\subsubsection*{The Dependency Graph: $\Gamma_{hybrid}$}
From the synchronized ordered list of computations $\Gamma_{data}$, a dependency graph $\Gamma_{hybrid}$ is built. A dependency exists between two computations (including synchronizations) if and only if a data read has been written by a previous computation in $\Gamma_{data}$, and if the computation domains intersect. In other words, the only case where a data dependency exists but is not taken into account is when the computation domains do not intersect. Nodes of the dependency graph represent computations of $\Gamma_{data}$, while edges are dependencies between them. The dependency graph $\Gamma_{hybrid}$ is a directed acyclic graph (\emph{dag}). For example, the dependency dag $\Gamma_{hybrid}$ of the example of Figure~\ref{fig:mslex} is displayed in Figure~\ref{fig:hyb}.

@@

\subsubsection*{The Minimal Series-Parallel Graph: $\Gamma_{msp}$}
Once a dependency graph is built, many solutions can be used to build a parallel application, as for example dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. In this work a static scheduling of the dependency graph is built. To do so the dependency graph is transformed to a minimal series-parallel graph~\cite{Valdes:1979:RSP:800135.804393}. As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a dag is a minimal series-parallel graph if and only if a forbidden shape, called \emph{N-shape}, is not found in the graph. To transform $\Gamma_{hybrid}$ to the minimal series-parallel graph $\Gamma_{msp}$, a breadth first search algorithm is applied to build a complete bipartite graph between two levels of $\Gamma_{hybrid}$~\cite{Mitchell:2004:CMV:1082101.1082117}.

>>>>

As mentionned in the previous section, the Multi Stencil Language (MSL) is an agnostic descriptive language for multi-stencil simulations. Six main sections are required in a MSL description, as the six tuples of the formal definition of a multi-stencil program: the mesh description; the mesh entities description; the computation domains description and their dependencies; the data description; the time loop description; and finally, the computations description. Figure~\ref{fig:grammar} shows the grammar of MSL. Lines 1 to 9 define what is a MSL program with the different parts memntionned above. The remaining of this section describes those different parts of a MSL program and presents an example.
We can note that, as MSL is a descriptive language, almost all terminals $meshid$, $meshdom$, $compdomid$, $dataid$, $compid$ and $neighborid$, are string identifiers. Only $num$ is an integer terminal.

\subsubsection*{Mesh and mesh entities}
As illustrated in the first line of the grammar in Figure~\ref{fig:grammar}, a mesh is directly defined by a string identifier. Then, mesh entities are described as a list of identifiers (lines 2, 10 and 11 in Figure~\ref{fig:grammar}). In the example given in Figure~\ref{fig:mslex} a Cartesian mesh, called $cart$, and three mesh entities, called $cell$, $edgex$ and $edgey$, are defined.

\subsubsection*{Computation domains and their dependencies}
A computation domain can be seen as a subpart of a mesh entity which will be used during one or more computation. For each computation domain, the mesh entity from which it is built is indicated (lines 3-4 and 12-14 in Figure~\ref{fig:grammar}). By default two computation domains always intersect except when an independency relation is mentionned in the section \emph{independent} of the language (lines 5-6 and 15-17 in Figure~\ref{fig:grammar}). In the example of Figure~\ref{fig:mslex}, on lines 4 to 6, a whole computation domain is defined for each mesh entity, called $allcell$, $alledgex$ and $alledgey$, and on lines 7 and 8 two subparts of the mesh entity $edgex$ are also defined, called $part1edgex$ and $part2edgex$. Finally, line 10 means that $part1edgex$ and $part2edgex$ do not intersect.

\subsubsection*{Data and time}
A data is a quantity to simulate and is mapped onto a mesh entity, as illustrated in lines 7 and 18-20 of Figure~\ref{fig:grammar}. The $time$ section simply indicates a number of iterations to perform in the simulation. In the example of Figure~\ref{fig:mslex} are defined ten data and 500 iterations (lines 11-22). An extension could be to support a convergence criterion.

\subsubsection*{Computations description}
Finally, the language follows the two definitions of a stencil kernel and a local computation given in Section~\ref{sect:multistencil}. However, as MSL does not handle numerical expressions, the term $exp$ of Equations~\ref{eq:st} and~\ref{eq:loc} is not used in the language. The exact code of the computation as to be writtenin into the final ready-to-fill parallel pattern of the simulation, using external languages or interfaces (\cf Section~\ref{sect:component}). Moreover, to propose an intuitive syntax, information are given in a different order than in formal definitions.
Thus, the description of a computation, as indicated in lines 24 and 25 of Figure~\ref{fig:grammar}, consists in first the identifier of the data to compute ($w$ in the formal description), in brackets the identifier of the computation domain on which the computation is applied ($d$ in the formal definition), an identifier for the computation kernel, and the set of identifiers for the data read by the computation, with parenthesis. For each data read by the computation brackets indicate the stencil shape identifier used on the data ($R$ in the formal definition of a stencil). If the computation is local no brackets appear ($R_l$ in the formal definition). In the example of Figure~\ref{fig:mslex}, nine computations are described. For example, on line 24, the data $b$ is computed on the computation domain $allcell$ by the computation $c_0$ which read data $a$ without neighborhood shape.

@@

%----------------------------------------
\subsection{MSC transformation}
%----------------------------------------
The MSC transformation, which is a subpart of the overall compiler MSCAC, makes use of the ordered list of computations $\Gamma$, which can directly be extracted from the parser, to build a parallel representation of the computations of the overall multi-stencil program. This transformation phase of the compilation is divided in five different steps; it transforms $\Gamma$ to a series-parallel tree decomposition~\cite{Valdes:1979:RSP:800135.804393}. This transformation phase handles data parallelism through the first step, and task parallelism through the three next steps. This section gives an overview of the four steps of the MSC transformation; the detailed formalism and the associated algorithms of those steps are described in the research report~\cite{????}.

\subsubsection*{The Ordered List of Computations: $\Gamma$}
$\Gamma$ is directly obtained from the parser. Actually, the list of computations in the MSL program is already ordered: $\Gamma$ is a direct map of this list. In the example of Figure~\ref{fig:mslex}, $\Gamma = [c_0,c_1,c_2,c_3,c_4,c_5,c_6,c_7,c_8]$.

\subsubsection*{The Synchronized Ordered List of Computations: $\Gamma_{sync}$}
Considering a data parallelization of the simulation, the mesh and data of the simulation are splitted among available processors (or cores) and a single program is applied on each sub-part with additionnal synchronizations between processors. For example, as a stencil kernel accesses neighborhood values, values computed by another processor are needed and synchronized. 
From the ordered list of computations $\Gamma$, the needed synchronizations can be automatically added. A synchronization is needed each time a data read by a stencil computation has been written by a previous computation. In such a case, a computation is added before the stencil computation. This \emph{synchronization computation} read the data to synchronize, and write the same data. The computation domain of such a synchronization computation is the mesh entity on which the data is declared. As a result $\Gamma$ is transformed to a synchronized ordered list of computations $\Gamma_{sync}$.
For example, in Figure~\ref{fig:mslex}, the stencil computation $c_1$ read the data $b$ which has been written by the computation $c_0$. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed to the sublist $[c_0,sync_1,c_1]$ in $\Gamma_{sync}$. The new computation $sync_1$ read and write $b$ and is applied on the computation domain of $c_1$. As a result, a dependency is kept between $c_0$ and $sync_1$ and between $sync_1$, and $c_1$. The same transformation is performed for all stencils: $c_4$ and $c_8$. Thus, $\Gamma_{sync} = [c_0,sync_1,c_1,c_2,c_3,sync_4,c_4,c_5,c_6,c_7,sync_8,c_8]$.

\subsubsection*{The Dependency Graph: $\Gamma_{dep}$}
From the synchronized ordered list of computations $\Gamma_{sync}$, a dependency graph $\Gamma_{dep}$ is built. A dependency exists between two computations (including synchronizations) if and only if a data read has been written by a previous computation in $\Gamma_{sync}$, and if the computation domains intersect. In other words, the only case where a data dependency exists but is not taken into account is when the computation domains do not intersect. Nodes of the dependency graph represent computations of $\Gamma_{sync}$, while edges are dependencies between them. The dependency graph $\Gamma_{dep}$ is a directed acyclic graph (\emph{dag}). For example, the dependency dag $\Gamma_{dep}$ of the example of Figure~\ref{fig:mslex} is displayed in Figure~\ref{fig:hyb}.

@@

\subsubsection*{The Minimal Series-Parallel Graph: $\Gamma_{msp}$}
Once a dependency graph is built, many solutions can be used to build a parallel application, as for example dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. In this work a static scheduling of the dependency graph is built. To do so the dependency graph is transformed to a minimal series-parallel graph~\cite{Valdes:1979:RSP:800135.804393}. As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a dag is a minimal series-parallel graph if and only if a forbidden shape, called \emph{N-shape}, is not found in the dag. Thus, to transform $\Gamma_{dep}$ to the minimal series-parallel graph $\Gamma_{msp}$, an algorithm~\cite{Mitchell:2004:CMV:1082101.1082117} is applied to remove all \emph{N-shapes}.
